Distinct smell and taste disorder phenotype of post-acute COVID-19 sequelae

### 2022-06-02 paper/markdown/supplementary_material.Rmd
Major analysis update
Corresponds now to the analyses presented in the pre-print

---
title: "Hyposmia as a standalone persistent symptom of long COVID"
subtitle: "Supplementary Material"
author: "Health after COVID-19 in Tyrol study team"
date: "`r format(Sys.time(), '%Y-%m-%d')`"

output: 
  pdf_document:
    latex_engine: xelatex
bibliography: cov_biblio.bib
csl: american-medical-association.csl
header-includes:
  \usepackage{longtable}
  \usepackage{tabu}
  \usepackage{caption}
  \usepackage{makecell}
  \usepackage{pdflscape}
  \usepackage{array}
  \usepackage{booktabs}
  \usepackage{threeparttable}
  \usepackage{threeparttablex}
  \usepackage{wrapfig}
  \usepackage{multirow}
  \usepackage[normalem]{ulem}
  \usepackage{colortbl}
  \usepackage{xcolor}
  \usepackage{float} \floatplacement{figure}{H} \floatplacement{table}{H}
  \newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{S\arabic{table}}       \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}
---

@@

\newpage
\beginsupplement

# Supplementary Methods

## Study design and participants

In the current project, the data obtained in the 'Health After COVID-19 in Tyrol' (HACT) online survey [@Sahanic2021; @Hufner2021] and during acute COVID-19, 60-, 100- and 180-day post-COVID-19 follow-up of the CovILD study [@Sonnweber2021; @Sonnweber2020; @Rass2021] were analyzed.

The HACT study collective encompassed two independently recruited cohorts of ambulatory COVID-19 convalescents in the Tyrol state in Austria and the South Tyrol province in Italy. The study inclusion criteria were laboratory-confirmed SARS-CoV-2 infection (PCR or seropositivity), adult age (Austria: $\geq$ 16 years, Italy: $\geq$ 18 years) and residence in the study regions [@Sahanic2021; @Hufner2021]. The analysis inclusion criterion was a minimal observation time of 90 days between the positive SARS-CoV-2 test and the survey completion (__Figure 1__). A total of `r nrow(hyposmia$survey_at) + nrow(hyposmia$survey_it)` questionnaires was eligible for the current analysis (Austria: n = `r nrow(hyposmia$survey_at)`, Italy: `r nrow(hyposmia$survey_it)`). All participants gave digitally signed informed consent before the survey start. The study was approved by the institutional review boards of the Medical University of Innsbruck (Austria, approval
number: 1257/2020) and of the Autonomous Province of Bolzano - South Tyrol (Italy, 0150701). The HACT study was registered at ClinicalTrials.gov (NCT: NCT04661462)

The CovILD cohort included a longitudinally investigated multi-center collective of convalescents of ambulatory (WHO categories 1 - 3, n = `r unique(kinet$analysis_tbl$covild$A$ID) %>% length`), moderate (WHO categories 4 - 5, n = `r unique(kinet$analysis_tbl$covild$HM$ID) %>% length`) and severe COVID-19 (WHO categories 6 - 9, n = `r unique(kinet$analysis_tbl$covild$HS$ID) %>% length`) recruited at the Department of Internal Medicine II at the Medical University of Innsbruck (Austria), St. Vinzenz Hospital in Zams (Austria) and the acute rehabilitation facility in Muenster (Austria) (__Figure 1__) [@Sonnweber2021; @Sonnweber2020; @Rass2021]. The study was approved by the institutional review board at the Medical University of Innsbruck (approval number: 1103/2020), and registered at ClinicalTrials.gov (NCT04416100).

## Measures, definitions and data transformation

In the HACT study, `r length(globals$symptoms)` symptoms of COVID-19 were surveyed. The participants were asked to classify them as 'absent', 'present for 1 - 3 days', 'present for up to 1 week', 'present for up to 2 weeks', 'present for up to 3 months', 'present for up to 6 months' or 'present for more than 6 months'. Based on that responses, individual symptom duration was calculated [@Sahanic2021]. In the CovILD study, `r length(kinet$variables$covild[kinet$variables$covild != 'sympt_present'])` self-reported symptoms were surveyed at each follow-up visit [@Sonnweber2020]. For both studies, detailed characteristic of demography, socioeconomic status, medical history before COVID-19 as well as the disease course and recovery was recorded as described [@Sonnweber2021; @Sonnweber2020; @Rass2021; @Sahanic2021; @Hufner2021]. The full list of the variables available for the current analysis is provided in __Supplementary Table \@ref(tab:tab-study-vars)__.

Self-reported symptoms were classified as acute when present in the fist weeks after clinical manifestation of COVID-19, persistent when lasting for at least 28 days and long-term persistent complaints when present for longer than 3 months (HACT) or present at the 100-day follow-up visit. For analysis, symptom counts and anti-SARS-CoV-2 S1/S2 antibody levels were stratified by quartiles [@Sahanic2021; @Sonnweber2021]. Laboratory readouts of inflammation, iron metabolism and hematopoiesis in the CovILD cohort were longitudinally determined by the local hospital laboratory and binarized with standard cutoffs [@Sonnweber2021; @Sonnweber2020]. Lung function and lung CT abnormality assessment in the CovILD cohort were performed as described [@Sonnweber2020; @Sonnweber2021]. Assessment of hyposmia by the 16 point sniffing stick test at the 100-day visit in the CovILD collective was done essentially as described before [@Rass2021]. For variable stratification scheme, see __Supplementary Table \@ref(tab:tab-study-vars)__.

## Statistical analysis
### Data transformation, visualization, descriptive statistic, hypothesis testing and symptom co-occurrence

Data transformation, analysis and visualization was accomplished with R 4.0.5, _tidyverse_ environment and _cowplot_ package [@Wickham2019; @Wickham2016; @Wilke2019]. As most of the investigated numeric variables were non-normally distributed as assessed by Shapiro-Wilk test, non-parametric tests: Mann-Whitney U and Kruskal-Wallis tests were used to determine statistical significance between analysis groups. Statistical significance for differences in frequency of categorical variables between analysis groups, $\chi^2$ test was applied.

The concordance of sniffing stick test-measured and self-reported hyposmia, was investigated with the Cohen's $\kappa$ statistic (package _vcd_, function _Kappa()_) [@Fleiss1969]. Co-occurrence of self-reported hyposmia and other symptoms (re-coded as 0: absent, 1: present) was measured by cosine similarity coefficient and cosine distance (package _philentropy_, function _distance_) [@Drost2018].

Test and modeling p values for each analysis task and cohort were corrected for multiple testing with Benjamini-Hochberg method [@Benjamini1995].

### Kinetic modeling

Symptom recovery kinetics was modeled by mixed-effect logistic regression (fixed terms: first and second time term, random term: individual, packages _lme4_ and _lmerTest_) [@Bates2015; @Kuznetsova2017] with the following general model formula:

$$Symptom \sim \beta_0 + \beta_1 \times t + \beta_2 \times t^2$$

where $t$ stands for time since symptom onset, $\beta_1$ was interpreted as the estimate of linear recovery and $\beta_2$ was interpreted as a parabolic 'plateau' term [@Box1978]. Significance of the 2^nd^ and 1^st^ order model terms was assessed by step-wise likelihood ratio test versus the lower order model (function _anova()_). Significance of the _beta_ estimates was determined by Wald Z test.

### Definition of the long-term COVID-19 recovery phenotypes

Clusters of HACT study participants with COVID-19 complaints present for at least 3 months (Austria: n = `r nrow(pheno$analysis_tbl$survey_at)`, Italy: n = `r nrow(pheno$analysis_tbl$survey_it)`) were defined in respect to the most frequent persistent features (present in at least 25% of participants for at least 28 days[@Sudre2021; @Sahanic2021]: impaired concentration, tachypnea, hyposmia, hypogeusia, fatigue, tiredness and forgetfulness; re-coded as 0: absent, 1: present). The clustering procedure employed a combined self-organizing map (SOM, `r pheno$som_grid$xdim` $\times$ `r pheno$som_grid$ydim` hexagonal grid, gaussian neighborhood function, Jaccard distance between the symptoms) and hierarchical clustering (HCL, Ward D2 algorithm, manhattan distance between the SOM nodes) [@Kohonen1995; @Vesanto2000; @Vesanto1999; @Hufner2021]. The clustering was accomplished with tools provided by _kohonen_ package [@Wehrens2018] and home-developed functions (https://github.com/PiotrTymoszuk/SOM_tools). The optimal number of final clusters was determined by the bend of the within-sum-of-square curve. Graphical representation of the clustering procedure training and cluster dentition in presented in __Supplementary Figure \@ref(fig:fig-clustering)__,

### Univariable modeling

To identify factors correlating with the risk of persistent long-term self-reported hyposmia (3 months or longer in the HACT study or present at the 100-day follow-up visit in the CovILD cohort) within the set of candidate parameters recorded during acute COVID-19 or, for the CovILD study, at the 60-day follow-up visit as well, a series of univariable logistic models was used. for the HACT study, the models were weighted for age and sex weights based on the age and sex distribution of the general COVID-19 convalescent population in Tyrol/Austria and Italy [@Sahanic2021; @IstitutoSuperiorediSanitaISS; @AustrianAgencyforHealthandFoodSafetyAGES]. To account for the effects of varying diagnosis - survey time in the HACT study, the stratified observation time variable was included as a confounder in each model [@Sahanic]. Odds ratio significance was determined by Wald Z test. Logistic modeling and visual model quality control was accomplished with home-developed wrappers (https://github.com/PiotrTymoszuk/lm_qc_tools) around base R modeling tools and diagnostic functions provided by _broom_ package [@Robinson2014].

# Data availability

The study data is available at reasonable request to the corresponding authors. The source code of the R analysis pipeline is available at GitHub (https://github.com/PiotrTymoszuk/hyposmia_analsis_pipeline).

@@

```{r tab-study-vars}

suppl_tables$study_vars[1, 1] %>% 
  mutate(`Variable in R code` = '') %>% 
  kable(caption = '\\textbf{Study variables.}\\\\Variable in R code: variable name in the analyis pipeline, Label: variable name used in plots and tables, Independent modeling variable: indicator if the variable was used for univariable hyposmia risk modeling.\\\\The table is available online.', 
        format = 'latex', 
        escape = T, 
        booktabs = T, 
        longtable = F, 
        col.names = '')

@@

```{r tab-kinetic}

suppl_tables$kinetic_modeling[1, 1] %>% 
  mutate(Study = '') %>% 
  kable(caption = '\\textbf{Results of symptom kinetic modeling.}\\\\OR: odds ratio with 95$\\%$ confidence interval, Significance OR: significance of the OR estimate obtained by Benjamini-Hochberg-corrected Wald Z test, Significance LRT: model term significance obtained by Benjamini-Hochberg-corrected likelihood ratio test.\\\\The table is available online.', 
        format = 'latex', 
        escape = T, 
        booktabs = T, 
        longtable = F, 
        col.names = '')

@@

```{r tab-similarity}

suppl_tables$hyposmia_similarity[1, 1] %>% 
  mutate(Symptom = '') %>% 
  kable(caption = '\\textbf{Values of cosine similarity coefficients for self-reported hyposmia and other COVID-19 symptoms.}\\\\The table is available online.', 
        format = 'latex', 
        escape = T, 
        booktabs = T, 
        longtable = F, 
        col.names = '')

@@

```{r tab-uni-modeling}

suppl_tables$uni_modeling[1, 1] %>% 
  mutate(Study = '') %>% 
  kable(caption = '\\textbf{Univariable modeling of long-term persistent hyposmia risk.}\\\\OR: odds ratio with 95$\\%$ confidence interval, Significance OR: significance of the OR estimate obtained by Benjamini-Hochberg-corrected Wald Z test.\\\\The table is available online.', 
        format = 'latex', 
        escape = T, 
        booktabs = T, 
        longtable = F, 
        col.names = '')

@@

```{r fig-beta, fig.width = mm_inch(suppl_figures$kinetic_beta$w), fig.height = mm_inch(suppl_figures$kinetic_beta$h), dpi = 600, fig.cap = 'Symptom recovery and chronicity term estimates obtained by kinetic modeling in the HACT study.'}

suppl_figures$kinetic_beta$plot

@@

__Supplementary Figure \@ref(fig:fig-beta). Symptom recovery and chronicity term estimates obtained by kinetic modeling in the HACT study.__

Symptom frequency during acute COVID-19 and recovery (0, 7, 14, 28 and 90 days post symptom onset) in the Austria (`r kinet$models$survey$survey_at %>% map_dbl(length) %>% paste_range`) and Italy (`r kinet$models$survey$survey_at %>% map_dbl(length) %>% paste_range`) HACT cohorts were modeled with mixed-effect logistic regression (__Supplementary Table \@ref(tab:tab-kinetic)__). Significance of the OR estimates was determined by Wald Z test. P values were corrected for multiple comparisons with Benjamini-Hochberg method. OR estimate values with 95% confidence intervals for the top 10 symptoms with the most significant second order terms as candidate long-term persistent features are presented as points and whiskers in the Forest plots. Dim.: diminished, imp.: impaired.

@@

```{r fig-low-chronicity, fig.width = mm_inch(suppl_figures$low_chronicity$w), fig.height = mm_inch(suppl_figures$low_chronicity$h), dpi = 600, fig.cap = 'Actual and predicted frequency of self-reported shivering and fever as symptoms characterized by an instant recovery in the HACT study.'}

suppl_figures$low_chronicity$plot

@@

__Supplementary Figure \@ref(fig:fig-low-chronicity). Actual and predicted frequency of self-reported shivering and fever as symptoms characterized by an instant recovery in the HACT study.__

Frequencies of self-reported shivering (__A__) and fever (__B__) during acute COVID-19 and recovery in the HACT study (0, 7, 14, 28 and 90 days post symptom onset) were modeled with mixed-effect logistic regression (__Supplementary Table \@ref(tab:tab-kinetic)__). Points and solid lines represent the actual frequencies. The model predictions are displayed as dashed lines. The numbers of complete observations are shown under the plots.

@@

```{r fig-high-chronicity, fig.width = mm_inch(suppl_figures$high_chronicity$w), fig.height = mm_inch(suppl_figures$high_chronicity$h), dpi = 600, fig.cap = 'Actual and predicted frequency of self-reported fatigue, tiredness and taste disorders as symptoms characterized by a protracted recovery in the HACT study.'}

suppl_figures$high_chronicity$plot

@@

__Supplementary Figure \@ref(fig:fig-high-chronicity). Actual and predicted frequency of self-reported fatigue, tiredness and taste disorders as symptoms characterized by a protracted recovery in the HACT study.__

>>>>

---
title: "Distinct smell and taste disorder phenotype of post-acute COVID-19 sequelae"
subtitle: "Supplementary Material"
author: "HACT and CovILD study teams"
date: "`r format(Sys.time(), '%Y-%m-%d')`"

output: 
  bookdown::word_document2:   
    reference_docx: ms_template.docx

bibliography: hyposmia_biblio.bib
csl: frontiers_medical.csl
header-includes:
  \usepackage{longtable}
  \usepackage{tabu}
  \usepackage{caption}
  \usepackage{makecell}
  \usepackage{pdflscape}
  \usepackage{array}
  \usepackage{booktabs}
  \usepackage{threeparttable}
  \usepackage{threeparttablex}
  \usepackage{wrapfig}
  \usepackage{multirow}
  \usepackage[normalem]{ulem}
  \usepackage{colortbl}
  \usepackage{xcolor}
  \usepackage{float} \floatplacement{figure}{H} \floatplacement{table}{H}
  \newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{S\arabic{table}}       \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}
---

@@

\beginsupplement

# Supplementary Methods
## Study procedures and variables

The complete list of study variables and stratification scheme is provided in __Supplementary Table S\@ref(tab:tab-hact-vars)__ for the HACT study and in __Supplementary Table S\@ref(tab:tab-covild-vars)__ for the CovILD study.

### COVID-19 symptoms

A total of `r length(globals$hact_symptoms)` was surveyed in the HACT study cohorts (__Supplementary Figure S\@ref(fig:fig-hact-freq)__ and __Supplementary Table S\@ref(tab:tab-hact-vars)__). The symptom duration was coded as follows: absent: 0 days, 1 - 3 days: 3 days, up to 1 week: 7 days, up to 2 weeks: 14 days, up to 4 weeks: 28 days, up to 3 months: 90 days, up to 6 months: 90 days, over 6 months: 90 days. Acute symptoms were defined as complaints present during the first 14 days after clinical onset of COVID-19. Long COVID symptoms were defined as complaints present for $\geq$ 28 days. PASC symptoms were defined as manifestations present for $\geq$ 90 days [@Venkatesan2021; @Shah2021].

In the CovILD study, a total of `r length(globals$covild_symptoms)` symptoms (reduced physical performance, hyposmia/anosmia, dyspnea, sleep problems, cough, fever, night sweating, gastrointestinal symptoms) were prospectively recorded with a standardized questionnaire at each of 60-, 100-, 180- and 360-day post COVID-19 follow-up (__Supplementary Figure S\@ref(fig:fig-covild-freq)__ and __Supplementary Table S\@ref(tab:tab-covild-vars)__). Acute COVID-19 symptoms were assessed retrospectively [@Sonnweber2020; @Sonnweber2022].

### Rating of physical recovery, mental health and quality of life in the HACT study

Self-perceived complete recovery, rehabilitation need and new medication since COVID-19 at the time of study participation were surveyed as single yes/no items. Percentage of physical performance loss as compared with the time before COVID-19 was rated with a 0 - 100% scale [@Hufner2022; @Sahanic2021]. Quality of life impairment (QoL) and overall mental health impairment (OMH) were rated with a 4 item Likert scale each (possible answers: “excellent,” “good,” “fair,” “poor,” scored: 0, 1, 2, and 3) [@Hufner2022; @Sahanic2021]. Anxiety (ANX) and depression (DPR) were assessed with the Patient Health Questionnaire (PHQ-4) [@Hufner2022; @Sahanic2021; @Lowe2010]. Psychosocial stress was scored with a modified 7 item PHQ stress module as described before [@Hufner2022; @Sahanic2021; @Grafe2004].

### Rating of hyposmia with sniffing stick test

Objective hyposmia at the 100-day and 360-day follow-up in the CovILD study participants was investigated with the 16-item sniffing stick test as described [@Rass2021]. Clinically relevant hyposmia was defined as $\leq$ 12 correct answers [@Rass2021; @Hummel2007]. In the analysis, participants with the complete answers concerning self-reported hyposmia and complete test results were included.

## Statistical analysis
### Data transformation, descriptive statistic

Data transformation and statistical analysis was done with R version 4.0.5 with _tidyverse_ data science environment [@Wickham2019]. Analysis results were visualized with _ggplot2_ [@Wickham2016], _cowplot_ [@Wilke2019] as well as in-house developed _ExDA_ (https://github.com/PiotrTymoszuk/ExDA) and _figur_ (https://github.com/PiotrTymoszuk/figur) packages.

Descriptive statistics including median with interquartile ranges and frequency of complete answers for numeric and categorical variables were calculated with base R functions and _ExDA_ package.

### Statistical hypothesis testing

Since multiple study variables were non-normally distributed as assessed by Shapiro-Wilk test and visual assessment of their distribution (quantile-quantile plots), statistical significance for differences in outcome numeric variables were assessed with Mann-Whitney U test by Wilcoxon r effect size statistic (two groups) or Kruskal-Wallis test with $\eta^2$ effect size statistic. Differences in frequency distribution for categorical outcome variables were assessed by $\chi^2$ test with Cramer V effect size statistic. Interrater assessment of self-reported and sniffing test hyposmia was accomplished with Cohen's $\kappa$ statistic [@Fleiss1969]. $\kappa$ significance ($\kappa\neq0$) was estimated with Wald Z test. P values were adjusted for multiple testing with Benjamini-Hochberg method [@Benjamini1995] separately for each analysis task and cohort. 
Packages _rstatix_ [@Kassambara2021], _vcd_ [@Meyer2021] and _Exda_ (https://github.com/PiotrTymoszuk/ExDA) were used for statistical hypothesis testing.

### Modeling of symptom recovery kinetic

To model recovery kinetics for binary symptom variables (0: absent, 1: present), second-order mixed-effect logistic (categorical features) modeling was applied (packages: _lme4_, _lmerTest_ and development package _kinet_ [https://github.com/PiotrTymoszuk/kinet]) [@Bates2015; @Kuznetsova2017; @Box1978]. Each model followed the general formula:

$$Response \sim time + time^2 + (1|individual)$$

where $(1|individual)$ indicates the random effect of the individual and $time$ and $time^2$ indicate the first- and second-order time effect terms. The first-order term estimate was interpreted as a measure of the recovery speed and the second-order term estimate was used to assess the plateau/rebound effect. Significance of the accuracy gain of the full second-order model compared with the nested null model was determined by likelihood ratio test (LRT) versus the nested first-order and null models, respectively. Likelihood ratio $\lambda$ statistic (full versus null model) was used as an effect size measure. 
Individuals from the HACT or CovILD studies with the complete longitudinal symptom record were included in the kinetic modeling tasks. Results of the kinetic modeling were adjusted for multiple comparisons with Benjamini-Hochberg method [@Benjamini1995].

## Symptom-symptom distances and multi-dimensional scaling

To assess co-occurrence or exclusivity of symptoms, simple matching distances between manifestations of each acute COVID-19, long COVID and PASC in the HACT study cohorts were calculated (package _scrime_ and development package _clustTools_ [https://github.com/PiotrTymoszuk/clustTools]) [@Boriah2008; @Schwender2018]. Subsequently, the distance matrix was subjected to multi-dimensional scaling (MDS, k = 2 dimensions, package _stats_, function _cmdscale()_). Association of specific symptoms was assessed by visual analysis of MDS coordinate plots.

### Apriori analysis of acute COVID-19, long COVID and PASC symptoms in the HACT study

Frequent combinations of symptoms of long COVID and PASC in the HACT study cohorts were identified with the apriori algorithm (package _arules_) [@Agrawal1993; @Hahsler2005] with the minimal support cutoff of 0.1, 2 - 10 item transaction length, confidence > 0.8 and lift > 2. The support statistic were used to estimate the symptom combination frequency. The confidence value was treated as an estimate of conditional probability of the symptom co-occurrence. The lift statistic was interpreted as a measure of the symptom dependence (lift = 1, symptoms are independent). The complete analysis results are presented in __Supplementary Tables S\@ref(tab:tab-apriori-long)__ and __S\@ref(tab:tab-apriori-pasc)__.

### Clustering analysis

COVID-19 recovery clusters of the training Autria (AT) HACT cohort participants in respect to symptom-specific recovery times (__Figure 1A__) were defined with the PAM (partitioning around medoids) algorithm and Euclidean distance statistic (packages _cluster_, _philentropy_ and development package _clustTools_ [https://github.com/PiotrTymoszuk/clustTools]) [@Drost2018; @Schubert2019]. The set of participants with the complete clustering variable set was included in the analysis. The symptom recovery times were not subjected to any type of pre-processing. The choice of the clustering procedure was motivated by the analysis of the clustering variance (ratio of the total between-cluster to total sum of squares) and clustering structure stability in 10-fold cross-validation (metric: rate of correct cluster assignment, cluster assignment predicted by k = 5 nearest neighbors label propagation algorithm, package _clustTools_) [@Lange2004; @Leng2014] for several clustering algorithms as presented in __Supplementary Figure S\@ref(fig:fig-clust-devel)A__. The optimal number of clusters was determined by the bend of the total within-cluster sum of squares curve (__Supplementary Figure S\@ref(fig:fig-clust-devel)B__, package _factoextra_) [@Kassambara2020]. 
Permutation importance of specific clustering variables was investigated by calculating difference in clustering variance (ratio of total between-cluster sum of squares to total sum of squares) between the initial clustering object and the clustering object with the given variable reshuffled at random (package _clustTools_). 
Assignment of the Italy HACT cohort participants to the recovery clusters was accomplished with k-nearest neighbors label propagation algorithm (k = 5) [@Leng2014]. The clustering efficacy in the training AT cohort and the test IT cohort measured by clustering variance statistic defined above was similar (AT: `r clustTools::var(part_clust$clust_north)$frac_var %>% signif(2)`, IT: `r clustTools::var(part_clust$clust_south)$frac_var %>% signif(2)`).

## Data and source code availability

The raw data files will be made available upon request. The entire analysis pipeline was published at https://github.com/PiotrTymoszuk/hyposmia_analsis_pipeline.

@@

```{r tab-hact-vars, tab.cap = 'HACT study variables.'}

suppl_tbl$hact_vars %>% 
  select(- Stratification) %>% 
  flextable %>% 
  width(1, width = 3.7, unit = 'cm') %>% 
  width(2, width = 3.9, unit = 'cm') %>% 
  width(3, width = 1.8, unit = 'cm')  %>% 
  width(4, width = 8.7, unit = 'cm') %>% 
  theme_vanilla %>% 
  font(part = 'all', fontname = 'Cambria') %>% 
  fontsize(size = 10, part = 'all')

@@

```{r tab-covild-vars, tab.cap = 'CovILD study variables.'}

suppl_tbl$covild_vars %>% 
  flextable %>% 
  width(1, width = 4.5, unit = 'cm') %>% 
  width(2, width = 4.5, unit = 'cm') %>% 
  width(3, width = 2, unit = 'cm')  %>% 
  theme_vanilla %>% 
  font(part = 'all', fontname = 'Cambria') %>% 
  fontsize(size = 10, part = 'all')

@@

```{r tab-apriori-long, tab.cap = 'Results of apriori analysis for long COVID-19 symptoms in the HACT study . Top 10 most frequent symptom combinations (highest support value) in the AT (Austria) and IT (Italy) cohort are presented. The full results are available as a supplementary Excel table.'}

suppl_tbl$apriori_long %>% 
  group_by(Cohort) %>% 
  top_n(10, Support) %>% 
  ungroup %>% 
  arrange(Cohort, Support) %>% 
  select(-Time) %>% 
  flextable %>% 
  width(1, width = 1.5, unit = 'cm') %>% 
  width(2, width = 9.5, unit = 'cm') %>% 
  width(3:5, width = 2.2, unit = 'cm')  %>% 
  theme_vanilla %>% 
  font(part = 'all', fontname = 'Cambria') %>% 
  fontsize(size = 10, part = 'all')

@@

```{r tab-apriori-pasc, tab.cap = 'Results of apriori analysis for PASC symptoms in the HACT study. Top 10 most frequent symptom combinations (highest support value) in the AT (Austria) and IT (Italy) cohort are presented. The full results are available as a supplementary Excel table.'}

suppl_tbl$apriori_pasc %>% 
  group_by(Cohort) %>% 
  top_n(10, Support) %>% 
  ungroup %>% 
  arrange(Cohort, Support) %>% 
  select(-Time) %>% 
  flextable %>% 
  width(1, width = 1.5, unit = 'cm') %>% 
  width(2, width = 9.5, unit = 'cm') %>% 
  width(3:5, width = 2.2, unit = 'cm')  %>% 
  theme_vanilla %>% 
  font(part = 'all', fontname = 'Cambria') %>% 
  fontsize(size = 10, part = 'all')

```{r tab-clust-base-at, tab.cap = 'Demographic and baseline clinical characteristic at the COVID-19 onset of the HACT study participants assigned to the recovery clusters, Austria (AT) cohort.'}

suppl_tbl$baseline_clusters_north %>% 
  flextable %>% 
  width(1, width = 3.5, unit = 'cm') %>% 
  width(2:4, width = 3.7, unit = 'cm') %>% 
  width(5, width = 2.6, unit = 'cm') %>% 
  width(6, width = 2.2, unit = 'cm') %>% 
  footnote(1, 5:6, value = as_paragraph('Categorical variables: \u03C7\u00B2 test with Cramer V effect size statistic. Numeric variables: Kruskal-Wallis test with \u03B7\u00B2 effect size statistic. P values corrected form multiple testing with Benjamini-Hochberg method.'), ref_symbols = 'a', part = 'header') %>% 
  footnote(3, 1, value = as_paragraph('BMI: body mass index, overweight > 25 kg/m\u00B2, obesity > 30 kg/m\u00B2,'), ref_symbols = 'b', part = 'body') %>% 
  footnote(15, 1, value = as_paragraph('Frequent respiratory infections, > 2 per year.'), ref_symbols = 'c', part = 'body') %>% 
  footnote(16, 1, value = as_paragraph('Frequent bacterial infections with antibiotic therapy, > 2 per year.'), ref_symbols = 'd', part = 'body') %>% 
  theme_vanilla %>%
  font(part = 'all', fontname = 'Cambria') %>% 
  fontsize(size = 10, part = 'all')

```{r tab-clust-base-it, tab.cap = 'Demographic and baseline clinical characteristic at the COVID-19 onset of the HACT study participants assigned to the recovery clusters, Italy (IT) cohort.'}

suppl_tbl$baseline_clusters_south %>% 
  flextable %>% 
  width(1, width = 3.5, unit = 'cm') %>% 
  width(2:4, width = 3.7, unit = 'cm') %>% 
  width(5, width = 2.6, unit = 'cm') %>% 
  width(6, width = 2.2, unit = 'cm') %>% 
  footnote(1, 5:6, value = as_paragraph('Categorical variables: \u03C7\u00B2 test with Cramer V effect size statistic. Numeric variables: Kruskal-Wallis test with \u03B7\u00B2 effect size statistic. P values corrected form multiple testing with Benjamini-Hochberg method.'), ref_symbols = 'a', part = 'header') %>% 
  footnote(3, 1, value = as_paragraph('BMI: body mass index, overweight > 25 kg/m\u00B2, obesity > 30 kg/m\u00B2,'), ref_symbols = 'b', part = 'body') %>% 
  footnote(15, 1, value = as_paragraph('Frequent respiratory infections, > 2 per year.'), ref_symbols = 'c', part = 'body') %>% 
  footnote(16, 1, value = as_paragraph('Frequent bacterial infections with antibiotic therapy, > 2 per year.'), ref_symbols = 'd', part = 'body') %>% 
  theme_vanilla %>%
  font(part = 'all', fontname = 'Cambria') %>% 
  fontsize(size = 10, part = 'all')

```{r tab-clust-reco-at, tab.cap = 'COVID-19 course and recovery in the HACT study participants assigned to the recovery clusters, Austria (AT) cohort.'}

suppl_tbl$cov_clusters_north %>% 
  flextable %>% 
  width(1, width = 3.5, unit = 'cm') %>% 
  width(2:4, width = 3.7, unit = 'cm') %>% 
  width(5, width = 2.6, unit = 'cm') %>% 
  width(6, width = 2.2, unit = 'cm') %>% 
  footnote(1, 5:6, value = as_paragraph('Categorical variables: \u03C7\u00B2 test with Cramer V effect size statistic. Numeric variables: Kruskal-Wallis test with \u03B7\u00B2 effect size statistic. P values corrected form multiple testing with Benjamini-Hochberg method.'), ref_symbols = 'a', part = 'header') %>% 
  footnote(3, 1, value = as_paragraph('Number of acute COVID-19 symptoms.'), ref_symbols = 'b', part = 'body') %>% 
  footnote(4, 1, value = as_paragraph('Number of long COVID symptoms.'), ref_symbols = 'c', part = 'body') %>% 
  footnote(7, 1, value = as_paragraph('Self-reported incomplete recovery.'), ref_symbols = 'd', part = 'body') %>% 
  footnote(11, 1, value = as_paragraph('ANX: anxiety.'), ref_symbols = 'e', part = 'body') %>% 
  footnote(12, 1, value = as_paragraph('DPR: depression.'), ref_symbols = 'f', part = 'body') %>% 
  footnote(14, 1, value = as_paragraph('OMH score: overall mental health impairment score'), ref_symbols = 'g', part = 'body') %>% 
  footnote(15, 1, value = as_paragraph('QoL score: quality of life impairment score'), ref_symbols = 'h', part = 'body') %>%
  theme_vanilla %>%
  font(part = 'all', fontname = 'Cambria') %>% 
  fontsize(size = 10, part = 'all')

```{r tab-clust-reco-it, tab.cap = 'COVID-19 course and recovery in the HACT study participants assigned to the recovery clusters, Italy (AT) cohort.'}

suppl_tbl$cov_clusters_south %>% 
  flextable %>% 
  width(1, width = 3.5, unit = 'cm') %>% 
  width(2:4, width = 3.7, unit = 'cm') %>% 
  width(5, width = 2.6, unit = 'cm') %>% 
  width(6, width = 2.2, unit = 'cm') %>% 
  footnote(1, 5:6, value = as_paragraph('Categorical variables: \u03C7\u00B2 test with Cramer V effect size statistic. Numeric variables: Kruskal-Wallis test with \u03B7\u00B2 effect size statistic. P values corrected form multiple testing with Benjamini-Hochberg method.'), ref_symbols = 'a', part = 'header') %>% 
  footnote(3, 1, value = as_paragraph('Number of acute COVID-19 symptoms.'), ref_symbols = 'b', part = 'body') %>% 
  footnote(4, 1, value = as_paragraph('Number of long COVID symptoms.'), ref_symbols = 'c', part = 'body') %>% 
  footnote(7, 1, value = as_paragraph('Self-reported incomplete recovery.'), ref_symbols = 'd', part = 'body') %>% 
  footnote(11, 1, value = as_paragraph('ANX: anxiety.'), ref_symbols = 'e', part = 'body') %>% 
  footnote(12, 1, value = as_paragraph('DPR: depression.'), ref_symbols = 'f', part = 'body') %>% 
  footnote(14, 1, value = as_paragraph('OMH score: overall mental health impairment score'), ref_symbols = 'g', part = 'body') %>% 
  footnote(15, 1, value = as_paragraph('QoL score: quality of life impairment score'), ref_symbols = 'h', part = 'body') %>%
  theme_vanilla %>%
  font(part = 'all', fontname = 'Cambria') %>% 
  fontsize(size = 10, part = 'all')

@@

```{r fig-hact-freq, fig.width = mm_inch(suppl$hact_freq$w), fig.height = mm_inch(suppl$hact_freq$h), fig.cap = 'Frequency of COVID-19 symptoms in acute COVID-19, long COVID and PASC in the HACT study.'}

suppl$hact_freq$plot

@@

__Supplementary Figure S\@ref(fig:fig-hact-freq). Frequency of COVID-19 symptoms in acute COVID-19, long COVID and PASC in the HACT study.__

Frequency of symptoms of acute COVID-19 (first 14 days post clinical onset), long COVID ($\geq$ 28 days) and PASC (post-acute sequelae of COVID-19, $\geq$ 90 days) in the AT (Austria, __A__) and IT (Italy, __B__) HACT study cohorts expressed as percentages of the cohort. Point size and color represents the percentage. Numbers of complete observations are indicated under the plot.

Dim. appetite: diminished appetite, Imp. concentration: impaired concentration, Imp. walk: impaired walk, Imp. FMS: impaired fine motor skills.

@@

```{r fig-hact-recovery, fig.width = mm_inch(suppl$hact_recovery$w), fig.height = mm_inch(suppl$hact_recovery$h), fig.cap = 'Kinetic of recovery from leading acute COVID-19 symptoms in the HACT study.'}

suppl$hact_recovery$plot

@@

__Supplementary Figure S\@ref(fig:fig-hact-recovery). Kinetic of recovery from leading acute COVID-19 symptoms in the HACT study.__

Percentages of individuals with fever (__A__), diminished appetite (__B__), joint pain (__C__), muscle pain (__D__), fatigue (__E__) and tachypnea (__F__) in the AT (Austria) and IT (Italy) HACT study cohorts at particular time points after clinical onset. Numbers of complete observations are indicated under the plots.

@@

```{r fig-covild-freq, fig.width = mm_inch(suppl$covild_freq$w), fig.height = mm_inch(suppl$covild_freq$h), fig.cap = 'Symptom frequency in ambulatory, moderate and severe COVID-19 subsets of the CovILD study.'}

suppl$covild_freq$plot

@@

__Supplementary Figure S\@ref(fig:fig-covild-freq). Symptom frequency in ambulatory, moderate and severe COVID-19 subsets of the CovILD study.__
